---
title: "Data Manipulation with `dplyr`"
author: "Elaine Yanxi Chen"
date: "2022-09-26"
output: github_document
---

```{r, echo = FALSE}
library(tidyverse)
```


Once you've imported data, you're going to need to do some cleaning up.

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = 0.6,
  out.width = "90"
)
```

## Overview {.tabset .tabset-pills}

### Learning Objectives

Clean and organize data using `dplyr` verbs and piping. 

```{r}
options(tibble.print_min = 3)

litters_data =
  read_csv("data/FAS_litters.csv")
litters_data = 
  janitor::clean_names(litters_data)

pups_data = read_csv("data/FAS_pups.csv")
pups_data = janitor::clean_names(pups_data)
```

### `select`

`select` is useful when we only want to select certain *columns* from the dataset for the analysis.

```{r}
select(litters_data, group, litter_number, gd0_weight, pups_born_alive)
```

Now we have 4 variables instead of 8. Here we didn't save it as a separate dataframe for now.

We could also specify a range of variable to include:

```{r}
select(litters_data, group:gd_of_birth)
```

This looks at the positioning of the variables inside a dataset. It selects every column within the range specified. Need to know the arrangement of variables in the dataset. 

You can also specify columns you would like to remove:

```{r}
select(litters_data, -pups_survive)
```

In addition, you could also rename variables as part of this process:

```{r}
select(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```
Here, we have a tibble with only these two variables.

If all you want to do is rename something, you can use `rename` instead of `select`. This will rename the variables we put, and keep everything else.

```{r}
rename(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```


There are some helper functions that go along with `select`; These three functions: `starts_with()`, `ends_with()`, and `contains()` are used very often, especially when there are variables named with suffixes or other standard patterns:

```{r}
select(litters_data, starts_with("gd"))
select(litters_data, ends_with("weight"))
```

Can read more using `?select_helpers`

Another frequently used function is `everything()`, which is handy for reorganizing columns without discarding anything (pull some variables to the front of the dataset before everything else):

```{r}
select(litters_data, litter_number, pups_survive, everything())
```

### `filter`

Select rows, certain observations, one individual. We use filter to achieve this. Select observations that meet certain criteria, using logical expressions. Use of comparison operators (`>`, `>=`, `<`, `<=`, `==`, and `!=`). Need to distinguish between assigning a value (a single equal sign and equal to.

Another operator to filter is in (`%in%`). Delete or remove missing data with `is.na()`. Now test some conditions, truth, go into the dataset; false, do not include. Logical operators and `&`, or `|`, or negate using `!`.

* `gd_of_birth == 20` to filter out all observations with gestational day of birth of 20
* `pups_born_alive >= 2`
* `pups_survive != 4`
* `!((pups_survive == 4) & (gd_of_birth == 20))`
* `group %in% c("Con7", "Con8")`
* `group == "Con7" & gd_of_birth == 20`

```{r}
filter(litters_data, gd_of_birth == 20)
```

```{r}
filter(litters_data, group == "Con7" & gd_of_birth == 20)
```

Be careful of dropping missing data, consider whether it makes sense. Can use filter or `drop_na` from the `tidyr` package

* `drop_na(litters_data)` will remove any row with a missing value; global drop of missing data.
* `drop_na(litters_data, wt_increase)` will remove rows for which `wt_increase` is missing.

### `mutate`

```{r}
mutate(litters_data,
       wt_gain = gd18_weight - gd0_weight,
       group = str_to_lower(group),
       # wt_gain_kg = wt_gain * 2.2
)
```


### `arrange`

Arrange the rows in your data according to the values in one or more columns, similar to sorting in SAS.

```{r}
head(arrange(litters_data, group, pups_born_alive), 10)
```

You can also sort in descending order if you'd like.

```{r}
head(arrange(litters_data, desc(group), pups_born_alive), 10)
```


### ` %>%`

Connect multiple functions or data manipulations together and output a single dataset at the end. 

```{r}
litters_data_raw = read_csv("data/FAS_litters.csv",
                            col_types = "ccddiiii")
litters_data_clean_names = 
  janitor::clean_names(litters_data_raw)
litters_data_selected_cols = select(litters_data_clean_names, -pups_survive)
litters_data_with_vars =
  mutate(
    litters_data_selected_cols,
    wt_gain = gd18_weight - gd0_weighht,
    group = str_to_lower(group))
litters_data_with_vars_without_missing =
  drop_na(litters_data_with_vars, wt_gain)
litters_data_with_vars_without_missing
```

Another way to do this before tidyverse is:

```{r}
litters_data_clean =
  drop_na(
    mutate(
      select(
        janitor::clean_names(
          read_csv("data/FAS_litters.csv", col_types = 
                     "ccddiiii")
        ),
        -pups_survive
      ),
      wt_gain = gd18_weight - gd0_weight,
      group = str_to_lower(group)
    ),
    wt_gain
  )
```

Now we use the pipe to achieve the same goal of obtaining a final dataset.

```{r}
litters_data = 
  read_csv("data/FAS_litters.csv", col_types = "ccddiiii") %>% 
  janitor::clean_names() %>% 
  select(-pups_survive) %>% 
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)) %>% 
  drop_na(wt_gain)
```

