---
title: "Data Manipulation with `dplyr`"
author: "Elaine Yanxi Chen"
date: "2022-09-26"
output: github_document
---

```{r, echo = FALSE}
library(tidyverse)
```


Once you've imported data, you're going to need to do some cleaning up.

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = 0.6,
  out.width = "90"
)
```

## Overview {.tabset .tabset-pills}

### Learning Objectives

Clean and organize data using `dplyr` verbs and piping. 

```{r}
options(tibble.print_min = 3)

litters_data =
  read_csv("data/FAS_litters.csv")
litters_data = 
  janitor::clean_names(litters_data)

pups_data = read_csv("data/FAS_pups.csv")
pups_data = janitor::clean_names(pups_data)
```

### `select`

`select` is useful when we only want to select certain *columns* from the dataset for the analysis.

```{r}
select(litters_data, group, litter_number, gd0_weight, pups_born_alive)
```

Now we have 4 variables instead of 8. Here we didn't save it as a separate dataframe for now.

We could also specify a range of variable to include:

```{r}
select(litters_data, group:gd_of_birth)
```

This looks at the positioning of the variables inside a dataset. It selects every column within the range specified. Need to know the arrangement of variables in the dataset. 

You can also specify columns you would like to remove:

```{r}
select(litters_data, -pups_survive)
```

In addition, you could also rename variables as part of this process:

```{r}
select(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```

If all you want to do is rename something, you can use `rename` instead of `select`. This will rename the variables 