---
title: "Data Import"
output: github_document
---

```{r, echo = FALSE}
library(tidyverse)
```


I'm an R Markdown document! 

## Data Import: CSVs

Let's import data using the `readr` package.

```{r}
litters_df = read_csv("data/FAS_litters.csv")
litters_df = janitor::clean_names(litters_df)
```


Look at the data

```{r}
litters_df

head(litters_df)
tail(litters_df)
```


Helpful way to view the whole dataset, but don't have it in the rmd file because it creates a new window and creates problems when knit.

```{r,  eval = FALSE}
View(litters_df)
```


The `skimr` package gives a quick summary and overview of the entire dataset. 

```{r}
skimr::skim(litters_df)
```


`read_csv` options..

can treat random numbers as missing observations when data are coded weirdly

Information in the first couple of rows, can skip how many rows at the beginning of the spreadsheet

```{r, eval = FALSE}
read_csv("data/FAS_litters.csv", na = c("", "NA", 999, 88), skip = 2)
```

Update with the col_type argument since it only reads the first 1000 rows and decide what variable type that is, which could be wrong. 


## Other file formats 